---
title: "Parametric Estimation of 1-D Stochastic Differential Equation"
author: 
- A.C. Guidoum^[Department of Probabilities & Statistics, Faculty of Mathematics, University of Science and Technology Houari Boumediene, BP 32 El-Alia, U.S.T.H.B, Algeria, E-mail  (acguidoum@usthb.dz)] and K. Boukhetala^[Faculty of Mathematics, University of Science and Technology Houari Boumediene, BP 32 El-Alia, U.S.T.H.B, Algeria, E-mail (kboukhetala@usthb.dz)]
date: "`r Sys.Date()`"
output: 
  knitr:::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Parametric Estimation of 1-D Stochastic Differential Equation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{latin1}
---

```{r setup, echo = F, message = F, results = 'hide',screenshot.force=FALSE}
library(Sim.DiffProc)
library(knitr)
knitr::opts_chunk$set(comment="",prompt=TRUE, fig.show='hold', warning=FALSE, message=FALSE)
options(prompt="R> ",scipen=16,digits=5,warning=FALSE, message=FALSE,
        width = 70)
```

# The `fitsde()` function

The [Sim.DiffProc](https://CRAN.R-project.org/package=Sim.DiffProc) package implements pseudo-maximum likelihood via the `fitsde()` function. The interface and the output of the `fitsde()` function are made as similar as possible to those of the standard `mle` function in the `stats4` package of the basic R system. The main arguments to `fitsde` consist: 

- `data` a univariate time series (`ts` object).
- `drift` and `diffusion` indicate drift and diffusion coefficient of the SDE, is an `expression` of two variables `t`, `x` and `theta` names of the parameters, and must be nominated by a vector of `theta = (theta[1], theta[2],..., theta[p])` for reasons of symbolic derived in approximation methods.
- `start` must be specified as a named list, where the names of the elements of the list correspond to the names of the parameters as they appear in the `drift` and `diffusion` coefficient.
- The `pmle` argument must be a `character` string specifying the method to use, can be either: `"euler"` Euler pseudo-likelihood, `"ozaki"` Ozaki pseudo-likelihood, `"shoji"` Shoji pseudo-likelihood and `"kessler"` Kessler pseudo-likelihood.
- `optim.method` select the optimization method (`"L-BFGS-B"` is used by default), and further arguments to pass to `optim` function.
- `lower` and `upper` bounds on the variables for the `Brent` or `L-BFGS-B` method.


The functions of type `S3 method` (as similar of the standard `mle` function in the `stats4` package of the basic R system for the class `fitsde` are the following:

- `coef`: which extracts model coefficients from objects returned by `fitsde`.
- `vcov`: returns the variance-covariance matrix of the parameters of a fitted model objects.
- `logLik`: extract log-likelihood.
- `AIC`: calculating Akaike's Information Criterion for fitted model objects.
- `BIC`: calculating Schwarz's Bayesian Criterion for fitted model objects.
- `confint`: computes confidence intervals for one or more parameters in a fitted model objects.

The following we explain how to use this function to estimate a SDE with different approximation methods.

## Euler method

Consider a process solution of the general stochastic differential equation:
\begin{equation}\label{eq02}
  dX_{t}= f(X_{t},\underline{\theta}) dt + g(X_{t},\underline{\theta}) dW_{t}, \quad \quad t \geq 0 \, , X_{0} = x_{0},
\end{equation}
The Euler scheme produces the discretization ($\Delta t \rightarrow 0$):
\begin{equation*}
X_{t+\Delta t} - X_{t} = f(X_{t},\theta) \Delta t+ g(X_{t},\theta) (W_{t+\Delta t} -W_{t}),
\end{equation*}
The increments $X_{t+\Delta t} - X_{t}$ are then independent Gaussian random variables with mean: $\text{E}_{x} = f(X_{t},\theta)\Delta t$, and variance: $\text{V}_{x} = g^{2}(X_{t},\theta) \Delta t$. Therefore the transition density of the process can be written as:
\begin{equation*}
  p_{\theta}(t,y|x)=\frac{1}{\sqrt{2\pi t g^{2}(x,\theta)}} \exp\left(-\frac{\left(y-x-f(x,\theta)t\right)^2}{2tg^{2}(x,\theta)}\right)
\end{equation*}
Then the log-likelihood is:
\begin{equation}\label{eq08}
  h_{n}(\theta|X_{1},X_{2},\dots,X_{n})=-\frac{1}{2}\left(\sum_{i=1}^{n} \frac{(X_{i}-X_{i-1}-f(X_{i-1},\theta)\Delta)^2}{\sigma^2 \Delta t} + n \log(2\pi \sigma^2 \Delta t)\right)
\end{equation}

As an example, we consider the Chan-Karolyi-Longstaff-Sanders (CKLS) model:
\begin{equation}\label{eq09}
  dX_{t} = (\theta_{1}+\theta_{2} X_{t}) dt + \theta_{3} X_{t}^{\theta_{4}} dW_{t},\qquad X_{0}=2
\end{equation}
with $\theta_{1}=1$, $\theta_{2}=2$, $\theta_{3}=0.5$ and $\theta_{4}=0.3$. Before calling `fitsde`, we generate sampled data $X_{t_{i}}$, with $\Delta t =10^{-4}$, as following:

```{r}
f <- expression( (1+2*x) ) ; g <- expression( 0.5*x^0.3 )
sim    <- snssde1d(drift=f,diffusion=g,x0=2,N=10^4,Dt=10^-4)
mydata <- sim$X
```
we set the initial values for the optimizer as $\theta_{1}=\theta_{2}=\theta_{3}=\theta_{4}=1$, and we specify the coefficients drift and diffusion as expressions.
you can use the `upper` and `lower` limits of the search region used by the optimizer (here using the default method `"L-BFGS-B"`), and specifying the method to use with `pmle="euler"`.

```{r, message=FALSE, warning=FALSE}
fx <- expression( theta[1]+theta[2]*x ) ## drift coefficient of model
gx <- expression( theta[3]*x^theta[4] ) ## diffusion coefficient of model 
fitmod <- fitsde(data = mydata, drift = fx, diffusion = gx, start = list(theta1=1, theta2=1,
                theta3=1,theta4=1),pmle="euler")
fitmod
```
The estimated coefficients are extracted from the output object `fitmod` as follows:
```{r}
coef(fitmod)
```
We can use the `summary` function to produce result summaries of output object:
```{r}
summary(fitmod)
```
`vcov` for variance-covariance matrice, and extract log-likelihood by `logLik`:
```{r}
vcov(fitmod)
logLik(fitmod)
AIC(fitmod)
BIC(fitmod)
```
Computes confidence intervals for one or more parameters in a fitted SDE:
```{r}
confint(fitmod, level=0.95)
```

## Ozaki method

The second approach we present is the Ozaki method, and it works for homogeneous stochastic differential equations. Consider the stochastic differential equation:
\begin{equation}\label{eq10}
  dX_{t}= f(X_{t},\underline{\theta}) dt + \sigma dW_{t}, \quad \quad t \geq 0 \, , X_{0} = x_{0},
\end{equation}
where $\sigma$ is supposed to be constant. We just recall that the transition density for the Ozaki method is Gaussian, we have that: $X_{t+\Delta t}|X_{t} = x \sim \mathcal{N}(\text{E}_{x},\text{V}_{x})$, where:
\begin{align}\label{eq11}
  \text{E}_{x} &= x + \frac{f(x)}{\partial_{x} f(x)} \left( e^{\partial_{x} f(x) \Delta t} - 1 \right), \quad\text{and}\quad
  \text{V}_{x} &= \sigma^{2} \frac{e^{2K_{x} \Delta t} -1}{2K_{x}},
\end{align}
with:
\begin{equation*}
  K_{x} = \frac{1}{\Delta t} \log \left(1+\frac{f(x)}{x\partial_{x}f(x)}\left(e^{\partial_{x}f(x) \Delta t}-1\right) \right)
\end{equation*}
It is always possible to transform process $X_t$ with a constant diffusion coefficient using the Lamperti transform.

Now we consider the Vasicek model, with $f(x,\theta) = \theta_{1} (\theta_{2}- x)$ and constant
volatility $g(x,\theta) = \theta_{3}$,
\begin{equation}\label{eq12}
  dX_{t} = \theta_{1} (\theta_{2}- X_{t}) dt + \theta_{3} dW_{t},\qquad X_{0}=5
\end{equation}
with $\theta_{1}=3$, $\theta_{2}=2$ and $\theta_{3}=0.5$, we generate sampled data $X_{t_{i}}$, with $\Delta t =10^{-2}$, as following:

```{r}
f <- expression( 3*(2-x) ) ; g <- expression( 0.5 )
sim <- snssde1d(drift=f,diffusion=g,x0=5,Dt=0.01)
HWV <- sim$X
```

we set the initial values for the optimizer as $\theta_{1}=\theta_{2}=\theta_{3}=1$, and we specify the coefficients drift and diffusion as expressions.
Specifying the method to use with `pmle="ozaki"`, which can easily be implemented in R as follows:

```{r, message=FALSE, warning=FALSE}
fx <- expression( theta[1]*(theta[2]- x) ) ## drift coefficient of model 
gx <- expression( theta[3] )           ## diffusion coefficient of model 
fitmod <- fitsde(data=HWV,drift=fx,diffusion=gx,start = list(theta1=1,theta2=1,
                  theta3=1),pmle="ozaki")
summary(fitmod)

```
If you want to have confidence intervals $\theta_{1}$ and $\theta_{2}$ parameters only, using the command `confint` as following:

```{r}
confint(fitmod,parm=c("theta1","theta2"),level=0.95)
```


## Shoji-Ozaki method

An extension of the method to Ozaki the more general case where the drift is allowed to depend on the time variable $t$, and also the diffusion coefficient can be varied is the method Shoji and Ozaki. Consider the stochastic differential equation:
\begin{equation}\label{eq13}
  dX_{t}= f(t,X_{t},\underline{\theta}) dt + g(X_{t},\underline{\theta}) dW_{t}, \quad \quad t \geq 0 \, , X_{0} = x_{0},
\end{equation}
the transition density for the Shoji-Ozaki method is Gaussian, we have that: $X_{t+\Delta t}|X_{t} = x \sim \mathcal{N}\left(\mathrm{A}_{(t,x)}x,\mathrm{B}^{2}_{(t,x)}\right)$, where:
\begin{align}\label{eq14}
  \mathrm{A}_{(t,x)} &= 1+ \frac{f(t,x)}{x\mathrm{L}_{t}} \left(e^{\mathrm{L}_{t}\Delta t }-1\right)+\frac{\mathrm{M}_{t}}{x\mathrm{L}^{2}_{t}} \left(e^{\mathrm{L}_{t} \Delta t}-1-\mathrm{L}_{t}\Delta t\right), \\
  \mathrm{B}_{(t,x)} &= g(x) \sqrt{\frac{e^{2\mathrm{L}_{t} \Delta t}-1}{2\mathrm{L}_{t}}},
\end{align}
with:
\begin{equation*}
  \mathrm{L}_{t} = \partial_{x} f(t,x) \quad\text{and}\quad  \mathrm{M}_{t} = \frac{g^{2}(x)}{2} \partial_{xx} f(t,x)+ \partial_{t} f(t,x).
\end{equation*}

As an example, we consider the following model:
\begin{equation}\label{eq15}
  dX_{t} = a(t)X_{t} dt + \theta_{2}X_{t} dW_{t},\qquad X_{0}=10
\end{equation}
with: $a(t) = \theta_{1}t$, and we generate sampled data $X_{t_{i}}$, with $\theta_{1}=-2$, $\theta_{2}=0.2$ and time step $\Delta t =10^{-3}$, as following:

```{r}
f <- expression(-2*x*t) ; g <- expression(0.2*x)
sim <- snssde1d(drift=f,diffusion=g,N=1000,Dt=0.001,x0=10)
mydata <- sim$X
```
we set the initial values for the optimizer as $\theta_{1}=\theta_{2}=1$, and we specify the method to use with `pmle="shoji"`:

```{r, message=FALSE, warning=FALSE}
fx <- expression( theta[1]*x*t ) ## drift coefficient of model 
gx <- expression( theta[2]*x )   ## diffusion coefficient of model 
fitmod <- fitsde(data=mydata,drift=fx,diffusion=gx,start = list(theta1=1,
                 theta2=1),pmle="shoji",lower=c(-3,0),upper=c(-1,1))
summary(fitmod)
```



## Kessler method

Kessler (1997) proposed to use a higher-order Ito-Taylor expansion to approximate the mean and variance in a conditional Gaussian density. Consider the stochastic differential equation $dX_{t}= f(X_{t},\underline{\theta}) dt + g(X_{t},\underline{\theta}) dW_{t}$ , the transition density by Kessler method is: $X_{t+\Delta t}|X_{t} = x \sim \mathcal{N}\left(\text{E}_{x},\text{V}_{x}\right)$, where:
\begin{align}\label{eq16}
  \text{E}_{x} &= x + f(t,x) \Delta t+\left(f(t,x)\partial_{x}f(t,x) + \frac{1}{2} g^{2}(t,x) \partial_{xx}g(t,x)\right)\frac{(\Delta t)^2}{2}, \\
  \text{V}_{x} &= x^2 +(2f(t,x)x+g^{2}(t,x)) \Delta t +\bigg(2f(t,x)\left(\partial_{x}f(t,x)x+f(t,x)+g(t,x)\partial_{x}g(t,x)\right) \nonumber\\
         &\quad+g^{2}(t,x)\left(\partial_{xx}f(t,x)x+2\partial_{x}f(t,x)+\partial_{x}g^{2}(t,x)+g(t,x)\partial_{xx}g(t,x)\right)\bigg)\frac{(\Delta t)^2}{2}-\text{E}^{2}_{x}.
\end{align}


We consider the following Hull-White (extended Vasicek) model:
\begin{equation}\label{eq17}
  dX_{t} = a(t)(b(t)-X_{t}) dt + \sigma(t) dW_{t},\qquad X_{0}=2
\end{equation}
with: $a(t) = \theta_{1}t$ and $b(t)=\theta_{2}\sqrt{t}$, the volatility depends on time: $\sigma(t)=\theta_{3}t$. We generate sampled data of $X_t$, with $\theta_{1}=3$, $\theta_{2}=1$ and $\theta_{3}=0.3$, time step $\Delta t =10^{-3}$, as following:

```{r}
f <- expression(3*t*(sqrt(t)-x)) ; g <- expression(0.3*t)
sim <- snssde1d(drift=f,diffusion=g,M=1,N=1000,x0=2,Dt=0.001)
mydata <- sim$X
```
we set the initial values for the optimizer as $\theta_{1}=\theta_{2}=\theta_{3}=1$, and we specify the method to use with `pmle="kessler"`:

```{r, message=FALSE, warning=FALSE}
fx <- expression( theta[1]*t* ( theta[2]*sqrt(t) - x ) ) ## drift coefficient of model 
gx <- expression( theta[3]*t ) ## diffusion coefficient of model 
fitmod <- fitsde(data=mydata,drift=fx,diffusion=gx,start = list(theta1=1,
                  theta2=1,theta3=1),pmle="kessler")
summary(fitmod)
```


# The `fitsde()` in practice

## Model selection via AIC

Let the following models:
\begin{align*}
% \nonumber to remove numbering (before each equation)
  dX_{t} &= \theta_{1} X_{t} dt + \theta_{2} X_{t}^{\theta_{3}} dW_{t},             &\text{(true model)}\\
  dX_{t} &= (\theta_{1}+\theta_{2} X_{t}) dt + \theta_{3} X_{t}^{\theta_{4}} dW_{t},&\text{(competing model 1)}\\
  dX_{t} &= (\theta_{1}+\theta_{2} X_{t}) dt + \theta_{3} \sqrt{X_{t}} dW_{t},      &\text{(competing model 2)}\\
  dX_{t} &= \theta_{1} dt + \theta_{2} X_{t}^{\theta_{3}} dW_{t},                   &\text{(competing model 3)}
\end{align*}
We generate data from true model with parameters $\underline{\theta}=(2,0.3,0.5)$, initial value $X_{0}=2$ and $\Delta t =10^{-4}$, as following:

```{r}
f <- expression( 2*x )
g <- expression( 0.3*x^0.5 )
sim <- snssde1d(drift=f,diffusion=g,M=1,N=10000,x0=2,Dt=0.0001)
mydata <- sim$X
```
We test the performance of the AIC statistics for the four competing models, we can proceed as follows:

```{r, message=FALSE, warning=FALSE}
## True model
fx <- expression( theta[1]*x )
gx <- expression( theta[2]*x^theta[3] )
truemod <- fitsde(data=mydata,drift=fx,diffusion=gx,start = list(theta1=1,
                   theta2=1,theta3=1),pmle="euler")
## competing model 1
fx1 <- expression( theta[1]+theta[2]*x )
gx1 <- expression( theta[3]*x^theta[4] )
mod1 <- fitsde(data=mydata,drift=fx1,diffusion=gx1,start = list(theta1=1,
          theta2=1,theta3=1,theta4=1),pmle="euler")
## competing model 2
fx2 <- expression( theta[1]+theta[2]*x )
gx2 <- expression( theta[3]*sqrt(x) )
mod2 <- fitsde(data=mydata,drift=fx2,diffusion=gx2,start = list(theta1=1,
           theta2=1,theta3=1),pmle="euler")
## competing model 3
fx3 <- expression( theta[1] )
gx3 <- expression( theta[2]*x^theta[3] )
mod3 <- fitsde(data=mydata,drift=fx3,diffusion=gx3,start = list(theta1=1,
           theta2=1,theta3=1),pmle="euler")
## Computes AIC
AIC <- c(AIC(truemod),AIC(mod1),AIC(mod2),AIC(mod3))
Test <- data.frame(AIC,row.names = c("True mod","Comp mod1","Comp mod2","Comp mod3"))
Bestmod <- rownames(Test)[which.min(Test[,1])]
Bestmod
```
the estimates under the different models:

```{r}
Theta1 <- c(coef(truemod)[[1]],coef(mod1)[[1]],coef(mod2)[[1]],coef(mod3)[[1]])
Theta2 <- c(coef(truemod)[[2]],coef(mod1)[[2]],coef(mod2)[[2]],coef(mod3)[[2]])
Theta3 <- c(coef(truemod)[[3]],coef(mod1)[[3]],coef(mod2)[[3]],coef(mod3)[[3]])
Theta4 <- c("",round(coef(mod1)[[4]],5),"","")
Parms  <- data.frame(Theta1,Theta2,Theta3,Theta4,row.names = c("True mod",
                      "Comp mod1","Comp mod2","Comp mod3"))
Parms
```

## Application to real data

We make use of real data of the U.S. Interest Rates monthly form $06/1964$ to $12/1989$ (see Figure 1) available in package [Ecdat](https://cran.r-project.org/package=Ecdat), and we estimate the parameters $\underline{\theta}=(\theta_{1},\theta_{2},\theta_{3},\theta_{4})$ of CKLS model. These data have been analyzed by Brouste et all (2014) with [yuima](https://cran.r-project.org/package=yuima) package, here we confirm the results of the estimates by several approximation methods.

```{r 01,fig.env='figure*', fig.cap='  '}
data(Irates)
rates <- Irates[, "r1"]
X <- window(rates, start = 1964.471, end = 1989.333)
plot(X)
```

we can now use all previous methods by `fitsde` function to estimate the parameters of CKLS model as follows:

```{r, message=FALSE, warning=FALSE}
fx <- expression( theta[1]+theta[2]*x ) ## drift coefficient of CKLS model
gx <- expression( theta[3]*x^theta[4] ) ## diffusion coefficient of CKLS model
pmle <- eval(formals(fitsde.default)$pmle)
fitres <- lapply(1:4, function(i) fitsde(X,drift=fx,diffusion=gx,pmle=pmle[i],
                  start = list(theta1=1,theta2=1,theta3=1,theta4=1)))
Coef <- data.frame(do.call("cbind",lapply(1:4,function(i) coef(fitres[[i]]))))
Info <- data.frame(do.call("rbind",lapply(1:4,function(i) logLik(fitres[[i]]))),
                   do.call("rbind",lapply(1:4,function(i) AIC(fitres[[i]]))),
                   do.call("rbind",lapply(1:4,function(i) BIC(fitres[[i]]))),
                   row.names=pmle)
names(Coef) <- c(pmle)
names(Info) <- c("logLik","AIC","BIC")
Coef
Info
```

In Figure 2 we reports the sample mean of the solution of the CKLS model with the estimated parameters and real data, their empirical $95\%$ confidence bands (from the $2.5th$ to the $97.5th$ percentile), we can proceed as follows:

```{r 02,fig.env='figure*', fig.cap='  '}
f <- expression( (2.076-0.263*x) )
g <- expression( 0.130*x^1.451 )
mod <- snssde1d(drift=f,diffusion=g,x0=X[1],M=500, N=length(X),t0=1964.471, T=1989.333)
mod
plot(mod,type="n",ylim=c(0,30))
lines(X,col=4,lwd=2)
lines(time(mod),apply(mod$X,1,mean),col=2,lwd=2)
lines(time(mod),apply(mod$X,1,bconfint,level=0.95)[1,],col=5,lwd=2)
lines(time(mod),apply(mod$X,1,bconfint,level=0.95)[2,],col=5,lwd=2)
legend("topleft",c("real data","mean path",paste("bound of", 95,"% confidence")),inset = .01,col=c(4,2,5),lwd=2,cex=0.8)

```

# Further reading

1. [`snssdekd()` & `dsdekd()` & `rsdekd()`- Monte-Carlo Simulation and Analysis of Stochastic Differential Equations](snssde.html).
2. [`bridgesdekd()` & `dsdekd()` & `rsdekd()` - Constructs and Analysis of Bridges Stochastic Differential Equations](bridgesde.html).
3. [`fptsdekd()` & `dfptsdekd()` - Monte-Carlo Simulation and Kernel Density Estimation of First passage time](fptsde.html).
4. [`MCM.sde()` & `MEM.sde()` - Parallel Monte-Carlo and Moment Equations for SDEs](mcmsde.html).
5. [`TEX.sde()` - Converting Sim.DiffProc Objects to LaTeX](sdetotex.html).
6. [`fitsde()` - Parametric Estimation of 1-D Stochastic Differential Equation](fitsde.html).




# References

1. Brouste A, Fukasawa M, Hino H, Iacus SM, Kamatani K, Koike Y, Masuda H, Nomura R,Ogihara T, Shimuzu Y, Uchida M, Yoshida N (2014). The YUIMA Project: A ComputationalFramework for Simulation and Inference of Stochastic Differential Equations." Journal of Statistical Software, 57(4), 1-51. URL http://www.jstatsoft.org/v57/i04.

2. Guidoum AC, Boukhetala K (2018). Sim.DiffProc: Simulation of Diffusion Processes. R package version 4.2, URL https://cran.r-project.org/package=Sim.DiffProc.

3. Iacus SM (2008). Simulation and Inference for Stochastic Differential Equations: With R Examples. Springer-Verlag, New York.

